### 2024-07-06 doc/Assignment2.qmd
Deleted unneccesary code and data

---
title: "Assignment II Analysis"
author: "Melisa Mazaeva"
date: "2024-07-12"
format: html
---

# Introduction
This document presents the analysis performed in Assignment 1, adapted to the open science workflow. Note that intermediate check steps have been omitted.

# Methodology
## Data Loading
```{python}
import pandas as pd
import numpy as np

# Load the data
static_data = pd.read_csv('data/external/wscp_static.txt', delimiter='\t')
panel_data = pd.read_excel('data/external/wscp_panel.xlsx')

# Display first few rows of the datasets
print(static_data.head())
print(panel_data.head())

### 2024-07-06 code/python/Assignment1.py

@@ -1,0 +1,263 @@
+import pandas as pd
+import numpy as np 
+
+
+'''
+Step 1
+'''
+# Load the data from Worldscope files into pandas DataFrames
+static_data = pd.read_csv('/Users/melissam/Documents/Uni🏅🩷/SS 24🥇/708014 Accounting Reading Group/Assignment I/wscp_static.txt', delimiter='\t') # use '\t' due to tab-separation in the static file
+panel_data = pd.read_excel('/Users/melissam/Documents/Uni🏅🩷/SS 24🥇/708014 Accounting Reading Group/Assignment I/wscp_panel.xlsx')
+# Normalize column names to avoid issues caused by variation in column naming conventions
+static_data.columns = static_data.columns.str.strip().str.lower().str.replace(' ', '_')
+panel_data.columns = panel_data.columns.str.strip().str.lower().str.replace(' ', '_')
+
+# Merge data on 'isin' to include country information in panel_data
+merged_data = panel_data.merge(static_data[['isin', 'country']], on='isin', how='left')
+# Print the merged data to check. Now we have added a column on countries in panel data
+print("Step 1: Merged data sample on 'isin' from static and panel data:\n", merged_data.head(11))
+print()  # add this to get a blank line for better readability
+# To understand how many rows there are in the merged data
+num_rows1 = len(merged_data)
+print("Number of rows in merged data sample on 'isin' from static and panel data:", num_rows1)
+print()
+print()
+
+
+'''
+Step 2
+'''
+# Filter the merged data for US firms and years 1996-2015
+us_firms = merged_data[merged_data['country'] == 'UNITED STATES'] # create a new DataFrame that contains only US firms
+filtered_data = us_firms[(us_firms['year_'] >= 1996) & (us_firms['year_'] <= 2015)].copy() # further filter the us_firms DataFrame to keep the task's required year range
+# Print the filtered data to check
+print("Step 2: Filtered data sample on US firms and years 1996-2015:\n", filtered_data.head(22))
+print()
+# To understand how many rows were dropped out
+num_rows2 = len(filtered_data)
+print("Number of rows in filtered data sample on US firms and years 1996-2015:", num_rows2)
+print()
+print()
+
+
+'''
+Step 3
+'''
+# Identify Year 0 to Year 6 for each firm within the filtered data
+filtered_data['year_0'] = filtered_data.groupby('isin')['year_'].transform('min') # this column will hold the according “Year 0” for each firm
[…]

### 2024-07-06 code/python/do_analysis.py

@@ -1,133 +1,0 @@
-import pickle
-import pandas as pd
-import numpy as np
-from plotnine import ggplot, geom_boxplot, labs, aes, geom_bin2d, after_stat
-from panel_eda_helper_funcs import (prepare_correlation_table,
-                                    prepare_descriptive_table, PrepareRegressionTable,
-                                    escape_for_latex)
-from prepare_data import treat_outliers
-from theme_trr import theme_trr, scale_color_trr266_d, scale_fill_trr266_c
-
-from utils import setup_logging, read_config
-
-log = setup_logging()
-
-
-def main():
-    log.info('Performing main analysis...')
-    cfg = read_config('config/do_analysis_cfg.yaml')
-    smp = pd.read_csv(cfg['acc_sample'], dtype={'gvkey': str})
-
-    fig_boxplot_full = make_boxplot(smp)
-
-    smp_da = prep_smp_da(smp)
-
-    fig_boxplot_smp = make_boxplot(smp_da)
-
-    fig_scatter_md_dd = make_scatter_plot(
-        smp_da, 'mj_da', 'dd_da', 'Modified Jones DA', 'Dechow and Dichev DA'
-    )
-
-    fig_scatter_dd_lnta = make_scatter_plot(
-        smp_da, 'ln_ta', 'dd_da', 'ln(Total Assets)', 'Dechow and Dichev DA'
-    )
-
-    fig_scatter_dd_roa = make_scatter_plot(
-        smp_da, 'ebit_avgta', 'dd_da', 'Return on Assets', 'Dechow and Dichev DA'
-    )
-    fig_scatter_dd_salesgr = make_scatter_plot(
-        smp_da, 'sales_growth', 'dd_da', 'Sales Growth', 'Dechow and Dichev DA'
-    )
-
-    tab_desc_stat = prepare_descriptive_table(smp_da.drop(['fyear'], axis=1))
-
-    desc_info = {
-        'min_fyear': smp_da['fyear'].min(),
-        'max_fyear': smp_da['fyear'].max(),
-        'no_unique_firms': smp_da['gvkey'].nunique()
-    }
-
-    tab_corr = prepare_correlation_table(smp_da.drop(['fyear'], axis=1))
-
-    smp_da = smp_da.set_index(['gvkey', 'fyear'])
-
-    tab_regression = PrepareRegressionTable(
-        smp_da,
-        dvs=['mj_da', 'dd_da'],
-        idvs=[
[…]

### 2024-07-06 code/python/panel_eda_helper_funcs.py

@@ -1,274 +1,0 @@
-# Panel Exploratory Data Analysis (PanelEDA) helper functions
-import re
-import string
-import pandas as pd
-import numpy as np
-from scipy.stats import spearmanr, pearsonr
-from linearmodels import PanelOLS
-
-
-def prepare_descriptive_table(df, precision=3):
-    df = df.select_dtypes(include=np.number)
-    desc = df.describe().transpose()
-    desc.columns = ["N", "Mean", "Std. dev.",
-                    "Min.", r"25 \%", "Median", r"75 \%", "Max."]
-    desc = desc.astype({'N': 'int'})
-
-    desc = desc.style.format(
-        precision=precision, thousands=',').to_latex(convert_css=True)
-
-    desc = re.sub(r"_", r"\\_", desc,  0, re.MULTILINE).split('\n')[0:-1]
-    desc[0] = re.sub(r"}{", r"}[t]{", desc[0])
-
-    desc = [r'\begin{table}', r'\caption{Descriptive Statistics}',
-            r'\centering', desc[0], r'\toprule', desc[1], r'\midrule'] + desc[2:-1] + [r'\bottomrule', desc[-1], r'\end{table}']
-    return '\n'.join(desc)+'\n'
-
-
-def _highlight_sig_corr(pval_tab):
-    return pval_tab.map(lambda y: 'font-weight:bold'.format() if y else '')
-
-
-def prepare_correlation_table(df, pval=0.05):
-    df = df.select_dtypes(include=np.number)
-    _spe, pval_tab = spearmanr(df)
-
-    letters = list(string.ascii_uppercase[0:len(df.columns)])
-    ind = [f'{letter}: {col}' for col, letter in zip(df.columns, letters)]
-
-    _spe = pd.DataFrame(_spe, columns=letters, index=ind)
-    pval_tab = pd.DataFrame(pval_tab, columns=letters, index=ind) < pval
-    pval_tab = pval_tab.astype(object)
-
-    for i in range(len(_spe)):
-        for j in range(len(_spe)):
-            if i == j:
-                _spe.iloc[i, j] = np.nan
-                pval_tab.iloc[i, j] = np.nan
-            if i < j:
-                person, pval_person = pearsonr(df.iloc[:, i], df.iloc[:, j])
-                _spe.iloc[i, j] = person
-                pval_tab.iloc[i, j] = pval_person < pval
-
[…]

### 2024-07-06 code/python/prepare_data.py

@@ -1,274 +1,0 @@
-# --- Header -------------------------------------------------------------------
-# Prepares the "Explore Discretionary Accruals" display
-#
-# (C) TRR 266 -  See LICENSE file for details
-# ------------------------------------------------------------------------------
-
-from itertools import product
-
-from statsmodels.formula.api import ols
-import numpy as np
-import pandas as pd
-
-from utils import read_config, setup_logging
-
-log = setup_logging()
-
-def main():
-    log.info("Preparing data for analysis ...")
-    cfg = read_config('config/prepare_data_cfg.yaml')
-
-    ff12 = pd.read_csv(cfg['fama_french_12'], dtype=object)
-    ff48 = pd.read_csv(cfg['fama_french_48'], dtype=object)
-
-    cstat_us_sample = pd.read_csv(cfg['cstat_us_sample'])
-    cstat_us_sample['gvkey'] = cstat_us_sample['gvkey'].astype(str)
-
-    us_base_sample = prep_us_base_sample(cstat_us_sample, ff12, ff48)
-
-    dup_obs = us_base_sample[us_base_sample.duplicated()]
-    assert dup_obs.shape[0] == 0, "Duplicate firm-year observations in Compustat data, stored in 'dup_obs'."
-
-    mj = estimate_mj_accruals(us_base_sample)
-    dd = estimate_dd_accruals(us_base_sample)
-
-    np.seterr(divide='ignore')  # because np.log(0) throws a warning
-    smp = prep_smp(us_base_sample, mj, dd)
-    np.seterr(divide='warn')
-
-    smp.to_csv(cfg['acc_sample'], index=False)
-
-    log.info("Preparing data for analysis ... Done!")
-
-
-def prep_us_base_sample(df, ff12, ff48):
-    df = df.astype({'sic': 'Int64', 'sich': 'Int64'})
-    df = (df
-          .assign(sic=lambda x: x["sich"].fillna(x["sic"]))
-          .query('indfmt == "INDL" & fic == "USA" & at.notna() & at > 0 & sale > 0')
-          .query('sic < 6000 | sic > 6999')
-          .query('sic.notna()')
-          .assign(sic=lambda x: x['sic'].astype(str).str.zfill(4))
-          .merge(ff48, how="left", on="sic")
-          .merge(ff12, how="left", on="sic")
[…]

### 2024-07-06 code/python/pull_wrds_data.py

@@ -1,77 +1,0 @@
-# --- Header -------------------------------------------------------------------
-# See LICENSE file for details
-#
-# This code pulls data from WRDS
-# ------------------------------------------------------------------------------
-import os
-from getpass import getpass
-import dotenv
-
-import pandas as pd
-from utils import read_config, setup_logging
-import wrds
-
-log = setup_logging()
-
-def main():
-    '''
-    Main function to pull data from WRDS.
-
-    This function reads the configuration file, gets the WRDS login credentials, and pulls the data from WRDS.
-
-    The data is then saved to a csv file.
-    '''
-    cfg = read_config('config/pull_data_cfg.yaml')
-    wrds_login = get_wrds_login()
-    wrds_us = pull_wrds_data(cfg, wrds_login)
-    wrds_us.to_csv(cfg['cstat_us_sample_save_path'], index=False)
-
-    
-def get_wrds_login():
-    '''
-    Gets the WRDS login credentials.
-    '''
-    if os.path.exists('secrets.env'):
-        dotenv.load_dotenv('secrets.env')
-        wrds_username = os.getenv('WRDS_USERNAME')
-        wrds_password = os.getenv('WRDS_PASSWORD')
-        return {'wrds_username': wrds_username, 'wrds_password': wrds_password}
-    else:
-        wrds_username = input('Please provide a wrds username: ')
-        wrds_password = getpass(
-            'Please provide a wrds password (it will not show as you type): ')
-        return {'wrds_username': wrds_username, 'wrds_password': wrds_password}
-    
-def pull_wrds_data(cfg, wrds_authentication):
-    '''
-    Pulls WRDS access data.
-    '''
-    db = wrds.Connection(
-        wrds_username=wrds_authentication['wrds_username'], wrds_password=wrds_authentication['wrds_password']
-    )
-
-    log.info('Logged on to WRDS ...')
-
-    dyn_var_str = ', '.join(cfg['dyn_vars'])
-    stat_var_str = ', '.join(cfg['stat_vars'])
-
-    log.info("Pulling dynamic Compustat data ... ")
-    wrds_us_dynamic = db.raw_sql(
-        f"SELECT {dyn_var_str} FROM comp.funda WHERE {cfg['cs_filter']}"
-    )
-    log.info("Pulling dynamic Compustat data ... Done!")
-
-    log.info("Pulling static Compustat data ... ")
[…]

### 2024-07-06 code/python/theme_trr.py

@@ -1,88 +1,0 @@
-from colorsys import hls_to_rgb, rgb_to_hls
-
-from plotnine import (element_blank, element_line, element_text,
-                      scale_color_gradient, scale_color_manual,
-                      scale_fill_gradient, scale_fill_manual, theme,
-                      theme_gray)
-
-
-def theme_trr(plot_type="default", legend=False, text_size=11, axis_y_horizontal=True):
-
-    plot_type = plot_type.lower()
-    valid_plots = ["bar", "box", "lollipop", "line",
-                   "scatter", "smoother", "default"]
-    if not plot_type in valid_plots:
-        raise ValueError(
-            "theme_trr: plot_type has to be 'default', 'bar', 'box', 'lollipop','line', 'scatter' or 'smoother'."
-        )
-
-    return (
-        theme_gray(base_family="sans") +
-        theme(
-            text=element_text(size=text_size),
-            axis_title_y=element_text(
-                angle=0) if axis_y_horizontal else element_text(angle=90),
-            axis_ticks=element_blank(),
-            axis_line=element_blank() if plot_type in [
-                "line", "scatter", "default"] else element_line(size=0.3, lineend="round"),
-            legend_position="top" if legend else "none",
-            legend_background=element_blank(),
-            legend_margin=1,
-            legend_key=element_blank(),
-            legend_direction="horizontal",
-            legend_box_background=element_blank(),
-            panel_background=element_blank(),
-            panel_grid=element_line(color="#EBEBEBFF", size=0.4),
-            panel_grid_minor_x=element_blank() if plot_type in [
-                "bar", "box", "line", "smoother", "default"] else None,
-            panel_grid_major_x=element_blank() if plot_type in [
-                "bar", "box", "line", "scatter", "default"] else None,
-            panel_grid_minor_y=element_blank() if plot_type in [
-                "lollipop", "scatter", "smoother"] else None,
-            panel_grid_major_y=element_blank() if plot_type == "lollipop" else None,
-            plot_background=element_blank(),
-            plot_title=element_text(weight="bold")
-        )
-    )
-
-
-col_trr266_petrol = "#1B8A8F"
[…]

### 2024-07-06 code/python/utils.py

@@ -1,21 +1,0 @@
-import logging
-import yaml
-
-def read_config(config_file):
-    '''
-    Reads the configuration yaml file.
-    '''
-    return yaml.safe_load(open(config_file, 'r'))
-
-
-def setup_logging():
-    '''
-    Sets up the logging configuration.
-    '''
-    logging.basicConfig(
-        level=logging.INFO,
-        format="%(asctime)s [%(levelname)s] %(message)s",
-        handlers=[logging.StreamHandler()],
-    )
-    log = logging.getLogger(__name__)
-    return log

### 2024-07-08 code/python/Assignment1.py
Adjusted script to save output files in the correct 'output' directory and fixed variable definition in Step 8

@@ -218,7 +218,7 @@ print("Median Residual Income by P/B Group and Relative Year:\n", median_residua
 print()
 
 # Save the results to an Excel file 
-median_residual_income.to_excel('median_residual_income.xlsx')
+median_residual_income.to_excel('output/median_residual_income.xlsx')
 
 
 '''
@@ -247,17 +247,17 @@ filtered_data['p/b_group'] = filtered_data.groupby('isin')['p/b_group'].ffill()
 median_pb_values = filtered_year_0_data.groupby('p/b_group', observed=True)['p/b'].median()
 
 # Group by P/B group and relative year, then calculate the median deflated residual income for each group and year
-median_residual_income = filtered_data.groupby(['p/b_group', 'relative_year'], observed=True)['deflated_residual_income'].median().unstack()
+median_residual_income_filtered = filtered_data.groupby(['p/b_group', 'relative_year'], observed=True)['deflated_residual_income'].median().unstack()
 
 # Formatting the results (decimal places)
-median_residual_income = median_residual_income.apply(lambda col: col.map(lambda x: f"{x:.3f}" if pd.notna(x) else "NaN"))
+median_residual_income_filtered = median_residual_income_filtered.apply(lambda col: col.map(lambda x: f"{x:.3f}" if pd.notna(x) else "NaN"))
 median_pb_values = median_pb_values.apply(lambda x: f"{x:.2f}")
 
 # Add the median P/B values column to the output table
-median_residual_income.insert(0, 'P/B', median_pb_values)
+median_residual_income_filtered.insert(0, 'P/B', median_pb_values)
 
 # Print the median residual income table
-print("Median Residual Income by P/B Group and Relative Year:\n", median_residual_income)
+print("Median Residual Income by P/B Group and Relative Year:\n", median_residual_income_filtered)
 
 # Save the results to an Excel file
-median_residual_income.to_excel('median_residual_income_filtered.xlsx')
+median_residual_income_filtered.to_excel('output/median_residual_income_filtered.xlsx')

### 2024-07-10 code/python/Assignment1.py
Updated data paths to relative paths for better portability

@@ -6,8 +6,8 @@ import numpy as np
 Step 1
 '''
 # Load the data from Worldscope files into pandas DataFrames
-static_data = pd.read_csv('/Users/melissam/Documents/Uni🏅🩷/SS 24🥇/708014 Accounting Reading Group/Assignment I/wscp_static.txt', delimiter='\t') # use '\t' due to tab-separation in the static file
-panel_data = pd.read_excel('/Users/melissam/Documents/Uni🏅🩷/SS 24🥇/708014 Accounting Reading Group/Assignment I/wscp_panel.xlsx')
+static_data = pd.read_csv('data/external/wscp_static.txt', delimiter='\t') # use '\t' due to tab-separation in the static file
+panel_data = pd.read_excel('data/external/wscp_panel.xlsx')
 # Normalize column names to avoid issues caused by variation in column naming conventions
 static_data.columns = static_data.columns.str.strip().str.lower().str.replace(' ', '_')
 panel_data.columns = panel_data.columns.str.strip().str.lower().str.replace(' ', '_')

### 2024-07-10 code/python/Assignment1_analyze_residual_income.py
Updated the code to save generated Excel files to the data/generated/ directory instead of the output/ directory

@@ -218,7 +218,7 @@ print("Median Residual Income by P/B Group and Relative Year:\n", median_residua
 print()
 
 # Save the results to an Excel file 
-median_residual_income.to_excel('output/median_residual_income.xlsx')
+median_residual_income.to_excel('data/generated/median_residual_income.xlsx')
 
 
 '''
@@ -260,4 +260,4 @@ median_residual_income_filtered.insert(0, 'P/B', median_pb_values)
 print("Median Residual Income by P/B Group and Relative Year:\n", median_residual_income_filtered)
 
 # Save the results to an Excel file
-median_residual_income_filtered.to_excel('output/median_residual_income_filtered.xlsx')
+median_residual_income_filtered.to_excel('data/generated/median_residual_income_filtered.xlsx')

### 2024-07-10 doc/paper.qmd
Deleted R markdown files (paper.Rmd and presentation.Rmd) to streamline the repository for Python and Quarto workflows only.

---
title: |
  | Exporting Discretionary Accruals: 
  | A Stub Project^[I acknowledge financial support from Deutsche Forschungsgemeinschaft-Project-ID 403041268 - TRR 266.] \vspace{1cm}
author: |
  | Joachim Gassen 
  | TRR 266 Accounting for Transparency
  | Humboldt-Universität zu Berlin 
  | gassen@wiwi.hu-berlin.de
date: today
date-format: MMM D, YYYY [\vspace{1cm}]
abstract: |
  | The Open Science movement promotes the accessibility and reusability of research. This repository has the objective to help researchers establishing such an collaboration-oriented workflow. It uses a toy project on discretionary accruals for demonstration. 
  | \vspace{6cm}
bibliography: references.bib
biblio-style: apsr
format:
  pdf:
    documentclass: article
    number-sections: true
    toc: false
fig_caption: yes
fontsize: 11pt
ident: yes
always_allow_html: yes
header-includes:
  - \usepackage[nolists]{endfloat}    
  - \usepackage{setspace}\doublespacing
  - \setlength{\parindent}{4em}
  - \setlength{\parskip}{0em}
  - \usepackage[hang,flushmargin]{footmisc}
  - \usepackage{caption} 
  - \captionsetup[table]{skip=24pt,font=bf}
  - \usepackage{array}
  - \usepackage{threeparttable}
  - \usepackage{adjustbox}
  - \usepackage{graphicx}
  - \usepackage{csquotes}
  - \usepackage[margin=1in]{geometry}
---

```{python}
#| echo: false
#| output: false

import pickle

with open('../output/results.pickle', 'rb') as f:
        results = pickle.load(f)

min_fyear, max_fyear, no_unique_firms = results['Desc information'].values()

def rename_var_to_label(table, italic=True):
    for var_name, var_label in results['Variable Names'].items():
        var_label = f'\\textit{{{var_label}}}' if italic else var_label
        table = table.replace(var_name, var_label)
    return table
```

\pagebreak

# Introduction

This is not a paper but a stub that is part of a [project template repository](https://github.com/trr266/treat). We developed this repository to provide a prototype for a reproducible and collaborative workflow. Several authors have discussed advantages of such workflows (@W_2017, @GGR_2018, @CFM_2019) and many scholars across fields have voiced the need for increased reproducibility in science (e.g., @I_2005, @GL_2014, @DPR_2017).

# Discretionary Accruals

To demonstrate our workflow, we explore discretionary accruals across the U.S. We calculate modified Jones and Dechow and Dichev type accruals and show their distributional properties. The main purpose of all this, however, is to provide a toy use case for [our project template directory](https://github.com/trr266/project_template) that contains all the code to obtain the data, run the analysis and prepare a paper as well as a presentation.

Table \ref{tab:tab-descriptives-sample} presents our data that is based on a simple WRDS pull of Compustat data with financial firms (SIC 6XXX) excluded. We require data to calculate all variables and this drastically reduces the sample size. Modified Jones discretionary accruals are calculated loosely based on @HN_2007 and Dechow and Dichev discretionary accruals are calculated based on (big surprise) @DD_2002. As you will see from @fig-boxplot_smp, discretionary accruals are very noisy constructs, even after limiting the sample to observations with complete data and winsorizing all data to the top and bottom percentile for each year. @fig-scatter_dd_lnta shows a very prominent heteroscedasticity of discretionary accruals with regards to size. While researchers have tried to address this problem, the distributional properties of these constructs significantly complicate the interpretation of discretionary accrual-related findings. Especially in high powered settings, the measurement error, being highly correlated with size, will tend to load on variables that are unrelated to the underlying economic construct but correlated with size. Table \ref{tab:corr} shows some correlations and Table \ref{tab:reg} shows some completely pointless regressions.

```{python}
#| label: fig-boxplot_smp
#| fig-cap: "Distribution of Discretionary Accruals over Time"
#| echo: false
#| fig-format: svg

results['fig_boxplot_smp']
```

```{python}
#| label: fig-scatter_dd_lnta
#| fig-cap: "Dechow and Dichev DA and Firm Size"
#| echo: false
#| fig-format: svg

results['fig_scatter_dd_lnta']
```

```{python}
#| echo: false
#| output: asis

lat_tab = rename_var_to_label(results['Descriptive Statistics']).split('\n')
lat_tab[4] = '\\\\[-1.8ex]\\hline \\hline \\\\[-1.8ex]'
lat_tab[6] = '\\hline\\\\[-1.8ex]'
lat_tab[-4] = '\\\\[-1.8ex]\\hline \\hline \\\\[-1.8ex]'

print('\n'.join([
    '\\begin{table}[!htbp]',  
    '\caption{\label{tab:descriptives}Descriptive Statistics}',
    '\label{tab:tab-descriptives-sample}',
    '\\begin{adjustbox}{width=\\textwidth}',
    '\\begin{threeparttable}',
    '\n'.join(lat_tab[3:-2]),
    '\\begin{tablenotes}[flushleft]',
    '\setlength{\labelsep}{0pt}',
    '\\footnotetext',
    f'\\item Note: The data is obtained from the Compustat U.S. as provided by WRDS. The sample covers the period {int(min_fyear)} to {int(max_fyear)} and {no_unique_firms:,} unique firms.',
    '\end{tablenotes}',
    '\end{threeparttable}',
    '\end{adjustbox}',
    '\\end{table}',
]))
```

```{python}
#| echo: false
#| output: asis

lat_tab = rename_var_to_label(results['Correlations'], False).split('\n')

lat_tab[2] = '\\\\[-0.9ex]\\hline \\hline \\\\[-0.9ex]'
lat_tab[4] = '\\hline\\\\[-1.8ex]'
lat_tab[12] = '\\\\[-0.9ex]\\hline \\hline \\\\[-0.9ex]'
lat_tab[14] = '\\begin{tablenotes}[flushleft]'

print('\n'.join([
    "\\begin{table}[!htbp]",
    "\centering",
    "\caption{\label{tab:CorrTable}Correlations}",
    '\label{tab:corr}',
    '\n'.join(lat_tab),
    "\end{table}"
]))
```

```{python}
#| echo: false
#| output: asis

print('\n'.join([
    '\\begin{table}[!htbp] \\centering',
    '\\caption{\\label{tab:reg}Regressions}', 
    '\label{tab:reg}',
    rename_var_to_label('\n'.join(results['Regression'].split('\n')[3:-1])),
    "\end{table}"
]))
```

# Conclusion

Isn't that wonderful? Discretionary accruals rock but what rocks even more is open science and a collaborative workflow. Clone or fork this repository to kickstart your own projects. If you do not like either R or Python, consider contributing code in your favorite statistical programming language to the repo. Thanks for reading and enjoy!

\pagebreak

\setcounter{table}{0}
\renewcommand{\thetable}{\arabic{table}}

# References {-}
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\noindent

### 2024-07-10 doc/paper_mine.qmd
Deleted R markdown files (paper.Rmd and presentation.Rmd) to streamline the repository for Python and Quarto workflows only.

---
title: |
  | Assignment II: 
  | Adopting an Open Science Workflow \vspace{1cm}
author: |
  | Melisa Mazaeva 
  | Accounting Reading Group – Summer 2024
  | Humboldt-Universität zu Berlin 
  | melisa.mazaeva@student.hu-berlin.de
date: today
date-format: MMM D, YYYY [\vspace{1cm}]
abstract: |
  | This project uses the TRR 266 Template for Reproducible Empirical Accounting Research (TREAT) and provides an infrastructure for open science oriented empirical projects. It has been adapted to analyze residual income based on grouped P/B ratios of U.S. public firms using external Worldscope data sets. It showcases a reproducible workflow integrating Python scripts and data analysis. The analysis is performed using a Python script that processes input data files from Wolrdscope and generates output documentation files with the results. This code base, adapted from TREAT, should give you an overview on how the template is supposed to be used for my specific project and how to structure a reproducible empirical project.
  | \vspace{6cm}
bibliography: references.bib
biblio-style: apsr
format:
  pdf:
    documentclass: article
    number-sections: true
    toc: false
fig_caption: yes
fontsize: 11pt
ident: yes
always_allow_html: yes
header-includes:
  - \usepackage[nolists]{endfloat}    
  - \usepackage{setspace}\doublespacing
  - \setlength{\parindent}{4em}
  - \setlength{\parskip}{0em}
  - \usepackage[hang,flushmargin]{footmisc}
  - \usepackage{caption} 
  - \captionsetup[table]{skip=24pt,font=bf}
  - \usepackage{array}
  - \usepackage{threeparttable}
  - \usepackage{adjustbox}
  - \usepackage{graphicx}
  - \usepackage{csquotes}
  - \usepackage[margin=1in]{geometry}
---

\pagebreak

# Introduction

The primary objective is to demonstrate a reproducible and collaborative research workflow by calculating and analyzing residual incomes and Price-to-Book (P/B) ratios. The analysis includes filtering relevant data, identifying key financial years, grouping firms based on their P/B ratios, and calculating delfated residual incomes for each firm/year.
Assignment I aims to replicate an empirical table that involves calculating the residual income of US firms over a defined period and examining the relationship between residual income and the Price-to-Book (P/B) ratio. The replication process includes data preparation, cleaning, normalization, and the application of statistical methods to compute and interpret financial metrics using Python. By following a step-by-step approach, I provide insights into the trends of residual earnings across different P/B groups, identical to trends presented by Penman (2013).
This document presents the analysis performed in Assignment 1, adapted to the open science workflow. Note that intermediate check steps have been omitted.

### 2024-07-10 doc/paper_mine.qmd
Prepared first successful draft of paper.pdf

@@
   - \usepackage{csquotes}
+  - \usepackage{indentfirst}  # Added this line to ensure the first paragraph is indented
   - \usepackage[margin=1in]{geometry}
 ---
+
+\pagebreak
+
+# List of Abbreviations
+\setlength{\parindent}{0em}
+
+**AMEX**: American Stock Exchange  
+**CAPM**: Capital Asset Pricing Model  
+**BVE**: Book Value of Equity  
+**IDE**: Integrated Development Environment  
+**MVE**: Market Value of Equity  
+**NYSE**: New York Stock Exchange  
+**P/B**: Price-to-Book  
+**UK**: United Kingdom  
+**US**: United States
+
+\setlength{\parindent}{4em}
 
@@
 The analysis includes filtering relevant data, identifying key financial years, grouping firms based on their P/B ratios, and calculating delfated residual incomes for each firm/year.
+
 Assignment I aims to replicate an empirical table that involves calculating the residual income of US firms over a defined period and examining the relationship between residual income and the Price-to-Book (P/B) ratio.
 The replication process includes data preparation, cleaning, normalization, and the application of statistical methods to compute and interpret financial metrics using Python.
-By following a step-by-step approach, I provide insights into the trends of residual earnings across different P/B groups, identical to trends presented by Penman (2013).
-This document presents the analysis performed in Assignment 1, adapted to the open science workflow. Note that intermediate check steps have been omitted.
+By following a step-by-step approach, I provide insights into the trends of residual earnings across different P/B groups, identical to trends presented by @Penman_2013.
+This document presents the analysis performed in Assignment I, adapted to the open science workflow.
+Note that intermediate check steps have been omitted.
+
+# Assumptions and Notes
+
+The aim of Assignment I is to replicate a specific empirical table that involves calculating the residual income of US firms over a defined period and examining the relationship between residual income and the Price-to-Book (P/B) ratio.
+The replication process involves data preparation, cleaning, and normalization, followed by the application of statistical methods to compute and interpret financial metrics.
+For Assignment I, I used the Python programming language to carry out the empirical analysis.
+Visual Studio Code was used as the Integrated Development Environment (IDE) for writing, debugging, and optimizing the Python code.
+
+## Assumptions
+1.
+A constant cost of equity is assumed to simplify the calculation of residual income across multiple years, making the analysis more straightforward and easier to interpret.
+This approach allows for a direct comparison of performance across firms and time periods without the increased complexity due to fluctuating rates.
+In contrary, estimating a dynamic cost of equity typically involves financial models such as the Capital Asset Pricing Model (CAPM), which would require additional data like risk-free rates and firm beta values that are not available.
+2.
+The level of cost of equity capital is assumed to be 8% (re = 0.08), a reasonable average estimate for the period 1996-2015 in the US (@Goedhart_Koller_Williams_2002; @Damodaran_2024).
+This is assumption reflects the return equity investors typically expect given the level of risk during that period.
+3. The entire Worldscope panel data provides the Book Value of Equity (BVE), Market Value of Equity (MVE), and net income as values for the end of the respective fiscal year.
+4.
+The beginning of Year 0 data is equivalent to the end of the previous year because the financial data recorded at the end of one year represents the starting values for the following year.
+Therefore, residual income is deflated by the BVE at the end of the year before Year 0 to ensure comparability across different firms and time periods.
+5.
+It is assumed that Penman (2013) restricted the possible P/B values to a range of 0 to 7 to exclude outliers.
+Hence, extreme P/B values that are negative and very high (greater than 7) are excluded to focus the analysis on firms with more stable and reasonable valuations, reducing the impact of outliers.
+
+By following the steps provided in Section 4 and adhering to the assumptions made, I successfully replicated the analysis and produced the required table.
+A thorough step-by-step approach helped to understand and verify the outputs.
+
+# Replication Steps
+
+```{python}
+#| label: step1
+#| echo: true
+#| output: true
+
+import pandas as pd
+import numpy as np
+import os
+
+# Paths to data files
+static_data_path = '/Users/melissam/Documents/GitHub/arg_trr266/data/external/wscp_static.txt' 
+panel_data_path = '/Users/melissam/Documents/GitHub/arg_trr266/data/external/wscp_panel.xlsx'
+
+# Check if files exist
+if not os.path.exists(static_data_path):
+    raise FileNotFoundError(f"{static_data_path} does not exist.")
+if not os.path.exists(panel_data_path):
+    raise FileNotFoundError(f"{panel_data_path} does not exist.")
+
+# Load the data from Worldscope files into pandas DataFrames
+static_data = pd.read_csv(static_data_path, delimiter='\t')  # use '\t' due to tab-separation in the static file
+panel_data = pd.read_excel(panel_data_path)
+
+# Normalize column names to avoid issues caused by variation in column naming conventions
+static_data.columns = static_data.columns.str.strip().str.lower().str.replace(' ', '_')
+panel_data.columns = panel_data.columns.str.strip().str.lower().str.replace(' ', '_')
+
+# Merge data on 'isin' to include country information in panel_data
+merged_data = panel_data.merge(static_data[['isin', 'country']], on='isin', how='left')
+
+# Print the merged data to check. Now we have added a column on countries in panel data
+print("Step 1: Merged data sample on 'isin' from static and panel data:\n", merged_data.head(11))
+print()  # add this to get a blank line for better readability
+# To understand how many rows there are in the merged data
+num_rows1 = len(merged_data)
+print("Number of rows in merged data sample on 'isin' from static and panel data:", num_rows1)
+print()
+print()
+```
+
+```{python}
+#| label: step2
+#| echo: true
+#| output: true
+
+# Filter the merged data for US firms and years 1996-2015
+us_firms = merged_data[merged_data['country'] == 'UNITED STATES']  # create a new DataFrame that contains only US firms
+filtered_data = us_firms[(us_firms['year_'] >= 1996) & (us_firms['year_'] <= 2015)].copy()  # further filter the us_firms DataFrame to keep the task's required year range
+
+# Print the filtered data to check
+print("Step 2: Filtered data sample on US firms and years 1996-2015:\n", filtered_data.head(22))
+print()
+
+# To understand how many rows were dropped out
+num_rows2 = len(filtered_data)
+print("Number of rows in filtered data sample on US firms and years 1996-2015:", num_rows2)
+print()
+print()
+```
+```{python}
+#| label: step3
+#| echo: true
+#| output: true
+
+# Identify Year 0 to Year 6 for each firm within the filtered data
+filtered_data['year_0'] = filtered_data.groupby('isin')['year_'].transform('min')  # this column will hold the according “Year 0” for each firm
+filtered_data['relative_year'] = filtered_data['year_'] - filtered_data['year_0']  # this column now indicates how many years have passed since the firm’s “Year 0”
+filtered_data = filtered_data[filtered_data['relative_year'] <= 6]  # filter the data to include only the first seven years (Year 0 to Year 6) for each firm
+
+# Print the filtered data to check if it's empty or has data
+print("Step 3: Filtered data sample after identifying the Year 0 and relative years:\n", filtered_data.head(22))
+print()
+
+# To understand how many rows were dropped out
+num_rows3 = len(filtered_data)
+print("Number of rows in filtered data sample after identifying the Year 0 and relative years:", num_rows3)
+print()
+print()
+```
+
+```{python}
+#| label: step4
+#| echo: true
+#| output: true
+
+# Calculate P/B Ratio for each firm in Year 0
+
+year_0_data = filtered_data[filtered_data['relative_year'] == 0].copy()  # create a copy of filtered_data DataFrame to contain only data for Year 0
+print("Step 4: Year 0 data:\n", year_0_data)
+print()
+
+year_0_data['bve'] = year_0_data['bve'].replace([0, np.inf, -np.inf], np.nan)  # replace zero and very small BVE values with NaN
+# To verify: display rows where 'bve' is NaN
+nan_bve_rows = year_0_data[year_0_data['bve'].isna()]
+print("Rows where 'bve' has been replaced with NaN:\n", nan_bve_rows)
+num_nan_bve_rows = len(nan_bve_rows)
+print("Number of rows where 'bve' has been replaced with NaN:", num_nan_bve_rows)
+print()
+
+# Use P/B Ratio formula to calculate it for each firm in Year 0
+year_0_data['p/b'] = year_0_data['mve'] / year_0_data['bve']
+
+# Before removing NaN P/B ratios
+num_rows_before = len(year_0_data)
+# Remove rows with NaN P/B ratios
+year_0_data = year_0_data.dropna(subset=['p/b'])
+# After removing NaN P/B ratios
+num_rows_after = len(year_0_data)
+
+# Calculate the number of dropped observations
+num_dropped = num_rows_before - num_rows_after
+print("Number of observations dropped due to NaN P/B ratios:", num_dropped)  # checked: the number of dropped observations is equal to number of dropped BVE observations
+print("Number of rows in Year 0 data after removing NaN P/B ratios:", num_rows_after)
+print("Year 0 data after calculating P/B ratio and dropping NaN values:\n", year_0_data)  # the DataFrame now contains new column p/b ratio
+print()
+print()
+```
+
+```{python}
+#| label: step5
+#| echo: true
+#| output: true
+
+# Form P/B groups
+
+# Sort firms based on their P/B ratios for Year 0 in descending order
+num_groups = 20  # define the number of groups
+year_0_data = year_0_data.sort_values('p/b', ascending=False).reset_index(drop=True)  # sort the year_0_data DataFrame in descending order based on the P/B ratios
+print("Step 5: Year 0 data sorted in descending order based on the P/B ratios:\n", year_0_data)  # check the correct order
+print()
+
+# Assign the groups so that the highest P/B ratios are in group 1 and lowest in group 20
+year_0_data['p/b_group'] = pd.qcut(year_0_data.index, num_groups, labels=range(1, num_groups + 1))
+print("Year 0 data updated with P/B groups:\n", year_0_data)  # checked: p/b ratios grouping aligns with previous output
+
+# Count the number of firms in each P/B group before merging
+pb_group_counts_pre_merge = year_0_data.groupby('p/b_group', observed=True)['isin'].nunique()
+print("Number of firms assigned to each P/B group before merging:\n", pb_group_counts_pre_merge)
+print()
+
+# Calculate median P/B value for each group
+median_pb_values = year_0_data.groupby('p/b_group', observed=True)['p/b'].median()
+# Display the median P/B values
+print("Median P/B values for each group:\n", median_pb_values)
+print()
+
+# Merge the P/B groups back to the filtered data for Year 0
+filtered_data = filtered_data.merge(year_0_data[['isin', 'p/b_group']], on='isin', how='left')
+# Fill missing P/B groups for years other than Year 0 with the same group as Year 0
+filtered_data['p/b_group'] = filtered_data.groupby('isin')['p/b_group'].ffill()
+
+# Display the combined filtered data after merging P/B groups
+print("Filtered data after merging P/B groups:\n", filtered_data.head(22))  # the DataFrame now contains new column p/b_group
+# Count the number of firms in each P/B group after merging
+pb_group_counts_post_merge = filtered_data.groupby('p/b_group', observed=True)['isin'].nunique()
+print("Number of firms assigned to each P/B group after merging:\n", pb_group_counts_post_merge)  # checked: same number as pre merging
+# Calculate the number of rows in the merged filtered data sample
+num_rows_merged_filtered = len(filtered_data)
+print("Number of rows in the merged filtered data sample:", num_rows_merged_filtered)  # checked: same number as pre merging
+print()
+```
+
+```{python}
+#| label: step6
+#| echo: true
+#| output: true
+
+# Calculate residual income for each P/B group for Years 0 to 6, deflated by the book value of equity at the end of the year before Year 0 (= beginning of year 0)
+
+cost_of_equity = 0.08  # assumption for cost of equity capital
+
+# Create a copy of the merged data from Step 1 to preserve all years of observations
+us_full_data = merged_data[merged_data['country'] == 'UNITED STATES'].copy()
+print("Step 6: Merged data (Step 1) filtered to US only, all years preserved:\n", us_full_data.head(22))
+print()
+
+# Identify Year 0 for each firm, ensuring Year 0 is between 1996 and 2015
+us_full_data['year_0'] = us_full_data.groupby('isin')['year_'].transform(lambda x: x[(x >= 1996) & (x <= 2015)].min())
+us_full_data = us_full_data[us_full_data['year_0'].notna()]  # remove rows where year_0 is NaN
+# Calculate the relative year and year_minus_1
+us_full_data['relative_year'] = us_full_data['year_'] - us_full_data['year_0']
+us_full_data['year_minus_1'] = us_full_data['year_0'] - 1
+
+# Convert years to integer
+us_full_data['year_0'] = us_full_data['year_0'].astype(int)
+us_full_data['relative_year'] = us_full_data['relative_year'].astype(int)
+us_full_data['year_minus_1'] = us_full_data['year_minus_1'].astype(int)
+# Print the data to verify the steps
+print("Data after identifying Year 0, relative year, and Year -1:\n", us_full_data.head(32))
+print()
+
+# Extract BVE for the year before Year 0 (relative_year = -1) for deflation
+bve_deflation = us_full_data[us_full_data['relative_year'] == -1][['isin', 'bve']].rename(columns={'bve': 'bve_deflation'})
+print(bve_deflation)
+# Merge the bve_deflation values back to the filtered data
+filtered_data = filtered_data.merge(bve_deflation, on='isin', how='left')
+print("Filtered data with bve_deflation:\n",filtered_data)  # new column on bve_deflation
+print()
+
+# Calculate the BVE of the previous year for each year (for residual income calculation)
+us_full_data['bve_prev'] = us_full_data.groupby('isin')['bve'].shift(1)
+# Merge bve_prev back to filtered_data
+filtered_data = filtered_data.merge(us_full_data[['isin', 'year_', 'bve_prev']], on=['isin', 'year_'], how='left')
+print("Filtered data with bve_prev:\n",filtered_data)  # new column on bve_prev
+print()
+
+# Function to calculate non-deflated residual income
+def calculate_residual_income(row):
+    return row['ninc'] - (cost_of_equity * row['bve_prev'])
+# Apply the function to calculate non-deflated residual income for each row
+filtered_data['residual_income'] = filtered_data.apply(calculate_residual_income, axis=1) 
+# Display the 'residual_income' values
+print("Residual income values:\n", filtered_data)  # new column on residual_income
+print()
+
+# Function to deflate residual income
+def deflate_residual_income(row):
+    if pd.notna(row['bve_deflation']) and row['bve_deflation'] != 0:
+        return row['residual_income'] / row['bve_deflation']
+    else:
+        return np.nan
+# Apply the function to deflate residual income for each row
+filtered_data['deflated_residual_income'] = filtered_data.apply(deflate_residual_income, axis=1)
+print("Deflated residual income values:\n", filtered_data)  # new column on deflated_residual_income
+print()
+```
+
+```{python}
+#| label: step7
+#| echo: true
+#| output: true
+
+# Output the replicated table
+
+# Group by P/B group and relative year, then calculate the median deflated residual income for each group and year
+median_residual_income = filtered_data.groupby(['p/b_group', 'relative_year'], observed=True)['deflated_residual_income'].median().unstack()
+
+# Formatting the results (decimal places)
+median_residual_income = median_residual_income.apply(lambda col: col.map(lambda x: f"{x:.3f}" if pd.notna(x) else "NaN"))
+median_pb_values = median_pb_values.apply(lambda x: f"{x:.2f}")
+
+# Add the median P/B values column to the output table
+median_residual_income.insert(0, 'P/B', median_pb_values)
+
+# Display the median residual income table
+print("Median Residual Income by P/B Group and Relative Year:\n", median_residual_income)
+print()
+```
+
+```{python}
+#| label: step8
+#| echo: true
+#| output: true
+
+# Step 8: Modification - exclude negative and very high P/B values
+
+# Filter out extreme P/B values that are negative and higher than 7
+filtered_year_0_data = year_0_data[(year_0_data['p/b'] > 0) & (year_0_data['p/b'] <= 7)]
+print("Filtered Year 0 data after excluding negative and very high P/B values:\n", filtered_year_0_data)
+print()
+
+# Recalculate the P/B groups based on the filtered data
+filtered_year_0_data = filtered_year_0_data.sort_values('p/b', ascending=False).reset_index(drop=True)
+filtered_year_0_data['p/b_group'] = pd.qcut(filtered_year_0_data.index, num_groups, labels=range(1, num_groups + 1))
+print("Filtered Year 0 data updated with P/B groups:\n", filtered_year_0_data)
+print()
+
+# Merge the P/B groups back to the filtered data for Year 0
+filtered_data = filtered_data.drop(columns=['p/b_group'])
+filtered_data = filtered_data.merge(filtered_year_0_data[['isin', 'p/b_group']], on='isin', how='left')
+filtered_data['p/b_group'] = filtered_data.groupby('isin')['p/b_group'].ffill()
+
+# Calculate median P/B values again
+median_pb_values = filtered_year_0_data.groupby('p/b_group', observed=True)['p/b'].median()
+
+# Group by P/B group and relative year, then calculate the median deflated residual income for each group and year
+median_residual_income_filtered = filtered_data.groupby(['p/b_group', 'relative_year'], observed=True)['deflated_residual_income'].median().unstack()
+
+# Formatting the results (decimal places)
+median_residual_income_filtered = median_residual_income_filtered.apply(lambda col: col.map(lambda x: f"{x:.3f}" if pd.notna(x) else "NaN"))
+median_pb_values = median_pb_values.apply(lambda x: f"{x:.2f}")
+
+# Add the median P/B values column to the output table
+median_residual_income_filtered.insert(0, 'P/B', median_pb_values)
+
+# Display the median residual income table
+print("Median Residual Income by P/B Group and Relative Year:\n", median_residual_income_filtered)
+```
+
+# Conclusion
+This project demonstrates the effectiveness of using an open science and collaborative workflow for analyzing residual incomes and Price-to-Book ratios of U.S.
+public firms.
+By following a step-by-step approach and using the TRR 266 Template for Reproducible Empirical Accounting Research, I was able to replicate an empirical table from @Penman_2013 and provide insights into the trends of residual earnings across different P/B groups, similar to @Penman_2013.
+
+The analysis revealed that high P/B groups initially exhibit higher residual earnings, which slowly decline over time, while low P/B groups consistently show negative residual earnings.
+Excluding extreme P/B values resulted in more stable and consistent residual earnings across groups and years.
+
+In the future, this repository can be cloned or forked (if made public) to kickstart further projects on residual income analysis. Thanks for reading and enjoy!
+
+\pagebreak
+
+\setcounter{table}{0}
+\renewcommand{\thetable}{\arabic{table}}
+
+# References {-}
+\setlength{\parindent}{-0.2in}
+\setlength{\leftskip}{0.2in}
+\setlength{\parskip}{8pt}
+\noindent

### 2024-07-10 doc/paper_mine.qmd
Correct data pathing in paper.qmd and according comment in README

@@
 import os
+import yaml
 
-# Paths to data files
-static_data_path = '/Users/melissam/Documents/GitHub/arg_trr266/data/external/wscp_static.txt' 
-panel_data_path = '/Users/melissam/Documents/GitHub/arg_trr266/data/external/wscp_panel.xlsx'
+# Function to get the path of the current file
+def get_current_path():
+    try:
+        return os.path.dirname(os.path.abspath(__file__))
+    except NameError:
+        return os.getcwd()
+
+# Define the relative path to the configuration file
+config_path = os.path.join(get_current_path(), '..', 'config', 'config.yaml')
+with open(config_path, 'r') as file:
+    config = yaml.safe_load(file)
+
+# Extract paths from the config file
+STATIC_DATA_PATH = os.path.join(get_current_path(), '..', config['data_paths']['static_data'])
+PANEL_DATA_PATH = os.path.join(get_current_path(), '..', config['data_paths']['panel_data'])
 
 # Check if files exist
-if not os.path.exists(static_data_path):
-    raise FileNotFoundError(f"{static_data_path} does not exist.")
-if not os.path.exists(panel_data_path):
-    raise FileNotFoundError(f"{panel_data_path} does not exist.")
+if not os.path.exists(STATIC_DATA_PATH):
+    raise FileNotFoundError(f"{STATIC_DATA_PATH} does not exist.")
+if not os.path.exists(PANEL_DATA_PATH):
+    raise FileNotFoundError(f"{PANEL_DATA_PATH} does not exist.")
 
 # Load the data from Worldscope files into pandas DataFrames
-static_data = pd.read_csv(static_data_path, delimiter='\t')  # use '\t' due to tab-separation in the static file
-panel_data = pd.read_excel(panel_data_path)
+static_data = pd.read_csv(STATIC_DATA_PATH, delimiter='\t')  # use '\t' due to tab-separation in the static file
+panel_data = pd.read_excel(PANEL_DATA_PATH)
 

### 2024-07-11 doc/paper_mine.qmd
adjustment

@@
 # Define the relative path to the configuration file
-config_path = os.path.join(get_current_path(), '..', 'config', 'config.yaml')
+current_path = get_current_path()
+project_root = os.path.abspath(os.path.join(current_path, '..'))
+config_path = os.path.join(project_root, 'config', 'config.yaml')
+
+# Load the configuration from YAML file
 with open(config_path, 'r') as file:
@@
 # Extract paths from the config file
-STATIC_DATA_PATH = os.path.join(get_current_path(), '..', config['data_paths']['static_data'])
-PANEL_DATA_PATH = os.path.join(get_current_path(), '..', config['data_paths']['panel_data'])
+STATIC_DATA_PATH = os.path.join(project_root, 'data', 'external', 'wscp_static.txt')
+PANEL_DATA_PATH = os.path.join(project_root, 'data', 'external', 'wscp_panel.xlsx')
 

### 2024-07-11 doc/paper_mine.qmd
Finalised paper.qmd

@@
 This document presents the analysis performed in Assignment I, adapted to the open science workflow.
-Note that intermediate check steps have been omitted.
+Note that intermediate check steps have been omitted for brevity.
+The full code, including intermediate print steps to verify outputs, is presented in `code/python/Assignment1_analyze_residual_income.py`.
 
@@
 # Replication Steps
+## Step 1: Data Loading and Merging
+Begin by loading the data from the provided Worldscope files.
+The static data contains information on unique ISIN IDs for each firm, as well as the corresponding firm name and country.
+The panel data provides the financial performance numbers of each ISIN.
+Merge the data on the ‘isin’ column to include country information in the panel data.
+This merging step is crucial for filtering the US firms specifically in further steps.
 
@@
 # Extract paths from the config file
-STATIC_DATA_PATH = os.path.join(project_root, 'data', 'external', 'wscp_static.txt')
-PANEL_DATA_PATH = os.path.join(project_root, 'data', 'external', 'wscp_panel.xlsx')
-
-# Check if files exist
-if not os.path.exists(STATIC_DATA_PATH):
-    raise FileNotFoundError(f"{STATIC_DATA_PATH} does not exist.")
-if not os.path.exists(PANEL_DATA_PATH):
-    raise FileNotFoundError(f"{PANEL_DATA_PATH} does not exist.")
+STATIC_DATA_PATH = os.path.join(project_root, 'data', 'external', 
+                                'wscp_static.txt')
+PANEL_DATA_PATH = os.path.join(project_root, 'data', 'external', 
+                               'wscp_panel.xlsx')
 
 # Load the data from Worldscope files into pandas DataFrames
-static_data = pd.read_csv(STATIC_DATA_PATH, delimiter='\t')  # use '\t' due to tab-separation in the static file
+static_data = pd.read_csv(STATIC_DATA_PATH, delimiter='\t')
 panel_data = pd.read_excel(PANEL_DATA_PATH)
 
-# Normalize column names to avoid issues caused by variation in column naming conventions
-static_data.columns = static_data.columns.str.strip().str.lower().str.replace(' ', '_')
-panel_data.columns = panel_data.columns.str.strip().str.lower().str.replace(' ', '_')
+# Normalize column names to avoid issues with column naming conventions
+static_data.columns = (
+    static_data.columns.str.strip()
+    .str.lower()
+    .str.replace(' ', '_')
+)
+
+panel_data.columns = (
+    panel_data.columns.str.strip()
+    .str.lower()
+    .str.replace(' ', '_')
+)
 
 # Merge data on 'isin' to include country information in panel_data
-merged_data = panel_data.merge(static_data[['isin', 'country']], on='isin', how='left')
-
-# Print the merged data to check. Now we have added a column on countries in panel data
-print("Step 1: Merged data sample on 'isin' from static and panel data:\n", merged_data.head(11))
-print()  # add this to get a blank line for better readability
-# To understand how many rows there are in the merged data
-num_rows1 = len(merged_data)
-print("Number of rows in merged data sample on 'isin' from static and panel data:", num_rows1)
-print()
-print()
+merged_data = panel_data.merge(
+    static_data[['isin', 'country']], 
+    on='isin', 
+    how='left'
+)
 ```
 
+## Step 2: Filter Data for US Firms and Relevant Years
+Filter the merged data to include only US firms (excluding UK and German firms) and the year range 1996-2015.
+This ensures that further calculations are based on the specific subset of data that matches the assignment’s criteria.
 ```{python}
@@
 
-# Filter the merged data for US firms and years 1996-2015
-us_firms = merged_data[merged_data['country'] == 'UNITED STATES']  # create a new DataFrame that contains only US firms
-filtered_data = us_firms[(us_firms['year_'] >= 1996) & (us_firms['year_'] <= 2015)].copy()  # further filter the us_firms DataFrame to keep the task's required year range
+us_firms = merged_data[merged_data['country'] == 'UNITED STATES']  
+filtered_data = us_firms[
+    (us_firms['year_'] >= 1996) & 
+    (us_firms['year_'] <= 2015)
+].copy() 
+```
 
-# Print the filtered data to check
-print("Step 2: Filtered data sample on US firms and years 1996-2015:\n", filtered_data.head(22))
-print()
+## Step 3: Identify Year 0 to Year 6 for Each Firm
+To calculate the P/B values and residual incomes for each firm/year, it is necessary to identify the base year (Year 0) and the subsequent relative years within the range 0 – 6 in the filtered data, as presented by Penman (2013).
+Define Year 0 as the earliest year in the dataset for each firm and display it in a new column ‘year_0’ that holds this value, making ‘year_0’ the same for each ISIN.
+Additionally, create the column ‘relative_year’, represented as the difference between the current year and Year 0.
+Since a firm may have more than 7 years of observations in the filtered data, limit the data to include only the first seven years (Year 0 to Year 6) for each ISIN to match the assignment’s requirements.
+Notably, the number of observations after merging Step 1 (298,120), filtering Step 2 (142,837), and identifying years in Step 3 (89,584) nearly halves with each step.
 
-# To understand how many rows were dropped out
-num_rows2 = len(filtered_data)
-print("Number of rows in filtered data sample on US firms and years 1996-2015:", num_rows2)
-print()
-print()
-```
 ```{python}
@@
 
-# Identify Year 0 to Year 6 for each firm within the filtered data
-filtered_data['year_0'] = filtered_data.groupby('isin')['year_'].transform('min')  # this column will hold the according “Year 0” for each firm
-filtered_data['relative_year'] = filtered_data['year_'] - filtered_data['year_0']  # this column now indicates how many years have passed since the firm’s “Year 0”
-filtered_data = filtered_data[filtered_data['relative_year'] <= 6]  # filter the data to include only the first seven years (Year 0 to Year 6) for each firm
+filtered_data['year_0'] = filtered_data.groupby('isin')['year_'].transform('min')  
+filtered_data['relative_year'] = filtered_data['year_'] - filtered_data['year_0']  
+filtered_data = filtered_data[filtered_data['relative_year'] <= 6]  
+```
 
-# Print the filtered data to check if it's empty or has data
-print("Step 3: Filtered data sample after identifying the Year 0 and relative years:\n", filtered_data.head(22))
-print()
-
-# To understand how many rows were dropped out
-num_rows3 = len(filtered_data)
-print("Number of rows in filtered data sample after identifying the Year 0 and relative years:", num_rows3)
-print()
-print()
-```
+## Step 4: Calculate P/B Ratio for Each Firm in Year 0
+According to Penman (2013), the residual earnings are calculated after P/B groups are formed in Year 0.
+Therefore, consider the firms’ financial performance data only for Year 0 to calculate the P/B ratio.
+Replace zero and infinite BVE values with NaN to drop invalid calculations
+(affecting 28 observations in total). Use the following formula to calculate the P/B ratio:
+$$
+P/B_t = \frac{\textit{Market value of equity}_t}{\textit{Book value of equity}_t}
+$$
 
@@
 
-# Calculate P/B Ratio for each firm in Year 0
+year_0_data = filtered_data[filtered_data['relative_year'] == 0].copy()
 
-year_0_data = filtered_data[filtered_data['relative_year'] == 0].copy()  # create a copy of filtered_data DataFrame to contain only data for Year 0
-print("Step 4: Year 0 data:\n", year_0_data)
-print()
-
-year_0_data['bve'] = year_0_data['bve'].replace([0, np.inf, -np.inf], np.nan)  # replace zero and very small BVE values with NaN
-# To verify: display rows where 'bve' is NaN
-nan_bve_rows = year_0_data[year_0_data['bve'].isna()]
-print("Rows where 'bve' has been replaced with NaN:\n", nan_bve_rows)
-num_nan_bve_rows = len(nan_bve_rows)
-print("Number of rows where 'bve' has been replaced with NaN:", num_nan_bve_rows)
-print()
+# replace zero and very small BVE values with NaN
+year_0_data['bve'] = year_0_data['bve'].replace([0, np.inf, -np.inf], np.nan)
 
@@
 
-# Before removing NaN P/B ratios
-num_rows_before = len(year_0_data)
 # Remove rows with NaN P/B ratios
 year_0_data = year_0_data.dropna(subset=['p/b'])
-# After removing NaN P/B ratios
-num_rows_after = len(year_0_data)
-
-# Calculate the number of dropped observations
-num_dropped = num_rows_before - num_rows_after
-print("Number of observations dropped due to NaN P/B ratios:", num_dropped)  # checked: the number of dropped observations is equal to number of dropped BVE observations
-print("Number of rows in Year 0 data after removing NaN P/B ratios:", num_rows_after)
-print("Year 0 data after calculating P/B ratio and dropping NaN values:\n", year_0_data)  # the DataFrame now contains new column p/b ratio
-print()
-print()
 ```
 
+## Step 5: Form P/B Groups
+Sort the firms based on their P/B ratios in descending order and divide them into 20 P/B groups, with the highest P/B ratios in group 1 and the lowest in group 20, as presented by Penman (2013).
+As a result, each P/B group contains approximately 900 firms, ensuring an equal division of the data set.
+Calculate the median P/B value for each group and merge the P/B group assignment information back into the filtered data, providing information on which P/B group each firm belongs to in each year.
 ```{python}
@@
 
-# Form P/B groups
-
 # Sort firms based on their P/B ratios for Year 0 in descending order
 num_groups = 20  # define the number of groups
-year_0_data = year_0_data.sort_values('p/b', ascending=False).reset_index(drop=True)  # sort the year_0_data DataFrame in descending order based on the P/B ratios
-print("Step 5: Year 0 data sorted in descending order based on the P/B ratios:\n", year_0_data)  # check the correct order
-print()
+year_0_data = year_0_data.sort_values(
+    'p/b', ascending=False
+).reset_index(drop=True)
 
-# Assign the groups so that the highest P/B ratios are in group 1 and lowest in group 20
-year_0_data['p/b_group'] = pd.qcut(year_0_data.index, num_groups, labels=range(1, num_groups + 1))
-print("Year 0 data updated with P/B groups:\n", year_0_data)  # checked: p/b ratios grouping aligns with previous output
-
-# Count the number of firms in each P/B group before merging
-pb_group_counts_pre_merge = year_0_data.groupby('p/b_group', observed=True)['isin'].nunique()
-print("Number of firms assigned to each P/B group before merging:\n", pb_group_counts_pre_merge)
-print()
+# Assign the groups so that the highest P/B ratios are in group 1
+year_0_data['p/b_group'] = pd.qcut(
+    year_0_data.index,
+    num_groups,
+    labels=range(1, num_groups + 1)
+)
 
@@
 median_pb_values = year_0_data.groupby('p/b_group', observed=True)['p/b'].median()
-# Display the median P/B values
-print("Median P/B values for each group:\n", median_pb_values)
-print()
 
 # Merge the P/B groups back to the filtered data for Year 0
-filtered_data = filtered_data.merge(year_0_data[['isin', 'p/b_group']], on='isin', how='left')
-# Fill missing P/B groups for years other than Year 0 with the same group as Year 0
+filtered_data = filtered_data.merge(
+    year_0_data[['isin', 'p/b_group']],
+    on='isin',
+    how='left'
+)
+
+# Fill missing P/B groups for years other than Year 0 with same group as Year 0
 filtered_data['p/b_group'] = filtered_data.groupby('isin')['p/b_group'].ffill()
+```
 
-# Display the combined filtered data after merging P/B groups
-print("Filtered data after merging P/B groups:\n", filtered_data.head(22))  # the DataFrame now contains new column p/b_group
-# Count the number of firms in each P/B group after merging
-pb_group_counts_post_merge = filtered_data.groupby('p/b_group', observed=True)['isin'].nunique()
-print("Number of firms assigned to each P/B group after merging:\n", pb_group_counts_post_merge)  # checked: same number as pre merging
-# Calculate the number of rows in the merged filtered data sample
-num_rows_merged_filtered = len(filtered_data)
-print("Number of rows in the merged filtered data sample:", num_rows_merged_filtered)  # checked: same number as pre merging
-print()
-```
+## Step 6: Calculate Residual Income
+This step involves calculating the residual income for each P/B group for Years 0 to 6, followed by deflation by the BVE at the beginning of Year 0.
+Assume a cost of equity capital of 8%, a reasonable assumption for US firms in the given time range.
+Based on Penman (2013), the residual income for each company/year should be deflated by the book value of equity at the beginning of Year 0.
+
+To identify whether the financial data from Worldscope contains the numbers as of the beginning or end of the fiscal year, a random sample check was conducted.
+For example, for ISIN AN8068571086 in its *2020 Annual Report* (2021), the values for total assets, stockholders’ equity, net income, and other financial metrics are measured at the end of each fiscal year [@Schlumberger_2021, p.
+20].
+Hence, assume that the entire Worldscope panel data provides the BVE, MVE, and net income as values for the end of the respective fiscal year.
+
+In case a company has observations before 1996, the filtered data set does not enable calculation of residual income in Year 0 (1996) due to missing value of BVE in Year -1, resulting into NaN.
+To avoid that, create a copy of the merged data from Step 1 to preserve all years of observations.
+Filter it to US companies and identify the year_minus_1 for each firm.
+Extract BVE for the year before Year 0 (relative_year = -1) for the deflation step and extract the BVE of the previous year for each year for the residual income calculation step.
+
+Calculate the non-deflated residual income for each firm/year using the following formula:
+$$
+\text{Residual income}_t = \text{Net income}_t - r_e \times \text{Book value of equity}_{t-1}
+$$
+
+Assume that the beginning of Year 0 is equivalent to the end of the previous year because the financial data recorded at the end of one year represents the starting values for the following year.
+According to Penman (2013), residual income is deflated by book value at the beginning of year 0.
+Hence, for deflating the residual income, use the book value of equity from the year before Year 0 (bve_deflation) for all years (Years 1 to 6).
+This means that the BVE used for deflation remains constant across all years, normalizing the residual income across different firms and time periods.
+For example, for Year 0 (1996), use the BVE at the end of 1995 to deflate the residual income.
+
+Calculate the deflated residual income for each firm/year using the following general formula:
+
+$$
+\text{Deflated residual income}_t = \frac{\text{Residual income}_t}{\text{Book value of equity}_{\text{Year 0 start}}}
+$$
 
@@
 
-# Calculate residual income for each P/B group for Years 0 to 6, deflated by the book value of equity at the end of the year before Year 0 (= beginning of year 0)
-
 cost_of_equity = 0.08  # assumption for cost of equity capital
 
-# Create a copy of the merged data from Step 1 to preserve all years of observations
+# Create a copy of the merged data from Step 1 to keep all years of observations
 us_full_data = merged_data[merged_data['country'] == 'UNITED STATES'].copy()
-print("Step 6: Merged data (Step 1) filtered to US only, all years preserved:\n", us_full_data.head(22))
-print()
 
 # Identify Year 0 for each firm, ensuring Year 0 is between 1996 and 2015
-us_full_data['year_0'] = us_full_data.groupby('isin')['year_'].transform(lambda x: x[(x >= 1996) & (x <= 2015)].min())
-us_full_data = us_full_data[us_full_data['year_0'].notna()]  # remove rows where year_0 is NaN
+us_full_data['year_0'] = us_full_data.groupby('isin')['year_'].transform(
+    lambda x: x[(x >= 1996) & (x <= 2015)].min()
+)
+us_full_data = us_full_data[us_full_data['year_0'].notna()] 
 # Calculate the relative year and year_minus_1
@@
 us_full_data['year_minus_1'] = us_full_data['year_minus_1'].astype(int)
-# Print the data to verify the steps
-print("Data after identifying Year 0, relative year, and Year -1:\n", us_full_data.head(32))
-print()
 
 # Extract BVE for the year before Year 0 (relative_year = -1) for deflation
-bve_deflation = us_full_data[us_full_data['relative_year'] == -1][['isin', 'bve']].rename(columns={'bve': 'bve_deflation'})
-print(bve_deflation)
+bve_deflation = (
+    us_full_data[us_full_data['relative_year'] == -1][['isin', 'bve']]
+    .rename(columns={'bve': 'bve_deflation'})
+)
+
 # Merge the bve_deflation values back to the filtered data
 filtered_data = filtered_data.merge(bve_deflation, on='isin', how='left')
-print("Filtered data with bve_deflation:\n",filtered_data)  # new column on bve_deflation
-print()
 
-# Calculate the BVE of the previous year for each year (for residual income calculation)
+# Calculate the BVE of the previous year for each year
 us_full_data['bve_prev'] = us_full_data.groupby('isin')['bve'].shift(1)
 # Merge bve_prev back to filtered_data
-filtered_data = filtered_data.merge(us_full_data[['isin', 'year_', 'bve_prev']], on=['isin', 'year_'], how='left')
-print("Filtered data with bve_prev:\n",filtered_data)  # new column on bve_prev
-print()
+filtered_data = filtered_data.merge(
+    us_full_data[['isin', 'year_', 'bve_prev']],
+    on=['isin', 'year_'],
+    how='left'
+)
 
@@
 # Apply the function to calculate non-deflated residual income for each row
-filtered_data['residual_income'] = filtered_data.apply(calculate_residual_income, axis=1) 
-# Display the 'residual_income' values
-print("Residual income values:\n", filtered_data)  # new column on residual_income
-print()
+filtered_data['residual_income'] = filtered_data.apply(
+    calculate_residual_income, axis=1
+)
 
@@
 # Apply the function to deflate residual income for each row
-filtered_data['deflated_residual_income'] = filtered_data.apply(deflate_residual_income, axis=1)
-print("Deflated residual income values:\n", filtered_data)  # new column on deflated_residual_income
-print()
+filtered_data['deflated_residual_income'] = filtered_data.apply(
+    deflate_residual_income, axis=1
+)
 ```
 
+## Step 7: Output the Replicated Table
+Calculate the median deflated residual income for each P/B group and relative year.
+Include the median P/B values in the output table.
+Ensure consistency with Penman (2013) by formatting the results to three decimal places for residual income and two decimal places for P/B values.
+The resulting table will be saved into an Excel file in the `data/generated` folder of the repository.
+
+The output table certainly indicates a wide range of P/B ratios among the firms, with some extreme values in high and low P/B groups.
+There is a noticeable trend where firms with very high or very low P/B ratios have less stable and more extreme residual earnings over time, while middle P/B groups seem to show more stable trends as presented by Penman (2013), suggesting a balanced reflection of expectations between market valuation and expected financial performance.
+
+The variation in trends can be explained by the difference in data sets and possible additional assumptions made by Penman (2013) that are not available.
+For example, the assignment data set includes all USA firms from the Worldscope database, providing a broader sample than Penman’s original study, which included only NYSE and AMEX firms.
+Furthermore, the assignment analysis focuses on the years 1996 to 2015, a more recent period with different economic conditions compared to the time frame used in Penman’s (2013) study.
 ```{python}
@@
 
-# Output the replicated table
-
-# Group by P/B group and relative year, then calculate the median deflated residual income for each group and year
-median_residual_income = filtered_data.groupby(['p/b_group', 'relative_year'], observed=True)['deflated_residual_income'].median().unstack()
+median_residual_income = filtered_data.groupby(
+    ['p/b_group', 'relative_year'], observed=True
+)['deflated_residual_income'].median().unstack()
 
 # Formatting the results (decimal places)
-median_residual_income = median_residual_income.apply(lambda col: col.map(lambda x: f"{x:.3f}" if pd.notna(x) else "NaN"))
+median_residual_income = median_residual_income.apply(
+    lambda col: col.map(
+        lambda x: f"{x:.3f}" if pd.notna(x) else "NaN"
+    )
+)
 median_pb_values = median_pb_values.apply(lambda x: f"{x:.2f}")
@@
 # Display the median residual income table
-print("Median Residual Income by P/B Group and Relative Year:\n", median_residual_income)
-print()
+print(
+    "Median Residual Income by P/B Group and Relative Year:\n", 
+    median_residual_income
+)
 ```
 
+## Step 8: Modification - Exclude Extreme P/B Values
+As indicated in Step 7, the output table certainly includes outliers.
+To replicate the more stable P/B values observed by Penman (2013), exclude extreme P/B values that are negative or higher than 7.
+Recalculate the P/B groups based on this filtered data.
+Merge the updated P/B group information back into the filtered data and repeat the calculation of median deflated residual income for each P/B group and relative year.
+Format the updated results with the required decimal places and save to an Excel file.
+The modified output of the replicated table, closely resembling Penman (2013), is presented in Table 2.
+
+The modified table resembles more the one of Penman (2013) by presenting similar trends.
+In particular, the exclusion of extreme P/B values results in more stable and consistent residual earnings across P/B groups and years.
+High P/B groups show higher residual earnings initially, which slowly decline over time, and low P/B groups have negative residual earnings that tend to fluctuate but still remain negative.
 ```{python}
@@
 
-# Step 8: Modification - exclude negative and very high P/B values
-
 # Filter out extreme P/B values that are negative and higher than 7
-filtered_year_0_data = year_0_data[(year_0_data['p/b'] > 0) & (year_0_data['p/b'] <= 7)]
-print("Filtered Year 0 data after excluding negative and very high P/B values:\n", filtered_year_0_data)
-print()
+filtered_year_0_data = year_0_data[
+    (year_0_data['p/b'] > 0) & (year_0_data['p/b'] <= 7)
+]
 
 # Recalculate the P/B groups based on the filtered data
-filtered_year_0_data = filtered_year_0_data.sort_values('p/b', ascending=False).reset_index(drop=True)
-filtered_year_0_data['p/b_group'] = pd.qcut(filtered_year_0_data.index, num_groups, labels=range(1, num_groups + 1))
-print("Filtered Year 0 data updated with P/B groups:\n", filtered_year_0_data)
-print()
+filtered_year_0_data = filtered_year_0_data.sort_values(
+    'p/b', ascending=False
+).reset_index(drop=True)
+
+filtered_year_0_data['p/b_group'] = pd.qcut(
+    filtered_year_0_data.index, num_groups, labels=range(1, num_groups + 1)
+)
 
@@
 filtered_data = filtered_data.drop(columns=['p/b_group'])
-filtered_data = filtered_data.merge(filtered_year_0_data[['isin', 'p/b_group']], on='isin', how='left')
+filtered_data = filtered_data.merge(
+    filtered_year_0_data[['isin', 'p/b_group']], 
+    on='isin', 
+    how='left'
+)
 filtered_data['p/b_group'] = filtered_data.groupby('isin')['p/b_group'].ffill()
@@
 # Calculate median P/B values again
-median_pb_values = filtered_year_0_data.groupby('p/b_group', observed=True)['p/b'].median()
+median_pb_values = filtered_year_0_data.groupby(
+    'p/b_group', 
+    observed=True
+)['p/b'].median()
 
-# Group by P/B group and relative year, then calculate the median deflated residual income for each group and year
-median_residual_income_filtered = filtered_data.groupby(['p/b_group', 'relative_year'], observed=True)['deflated_residual_income'].median().unstack()
+# Calculate the median deflated residual income for each group and year
+median_residual_income_filtered = filtered_data.groupby(
+    ['p/b_group', 'relative_year'], 
+    observed=True
+)['deflated_residual_income'].median().unstack()
 
 # Formatting the results (decimal places)
-median_residual_income_filtered = median_residual_income_filtered.apply(lambda col: col.map(lambda x: f"{x:.3f}" if pd.notna(x) else "NaN"))
+median_residual_income_filtered = median_residual_income_filtered.apply(
+    lambda col: col.map(lambda x: f"{x:.3f}" if pd.notna(x) else "NaN")
+)
 median_pb_values = median_pb_values.apply(lambda x: f"{x:.2f}")
@@
 # Display the median residual income table
-print("Median Residual Income by P/B Group and Relative Year:\n", median_residual_income_filtered)
+print(
+    "Median Residual Income by P/B Group and Relative Year:\n",
+    median_residual_income_filtered
+)
 ```

### 2024-07-11 doc/presentation_mine.qmd
Created documentation presentation

--- 2024-07-11 doc/paper_mine.qmd
@@
 ---
-title: |
-  | Assignment II: 
-  | Adopting an Open Science Workflow \vspace{1cm}
-author: |
-  | Melisa Mazaeva 
-  | Accounting Reading Group – Summer 2024
-  | Humboldt-Universität zu Berlin 
-  | melisa.mazaeva@student.hu-berlin.de
+title: "Assignment II: Adopting an Open Science Workflow"
+author: "Melisa Mazaeva"
+institute: "Accounting Reading Group – Summer 2024"
 date: today
-date-format: MMM D, YYYY [\vspace{1cm}]
-abstract: |
-  | This project uses the TRR 266 Template for Reproducible Empirical Accounting Research (TREAT) and provides an infrastructure for open science oriented empirical projects.
-It has been adapted to analyze residual income based on grouped P/B ratios of U.S.
-public firms using external Worldscope data sets.
-It showcases a reproducible workflow integrating Python scripts and data analysis.
-The analysis is performed using a Python script that processes input data files from Wolrdscope and generates output documentation files with the results.
-This code base, adapted from TREAT, should give you an overview on how the template is supposed to be used for my specific project and how to structure a reproducible empirical project.
-  | \vspace{6cm}
+fontsize: "9pt"
+pdf-engine: xelatex # pdflatex creates rastered fonts
+format: 
+  beamer:
+    slide-level: 3
+    number-sections: true
+    toc: false
+header-includes:
+- \usepackage{booktabs} 
+- \usepackage{threeparttable}
+- \usepackage{graphicx}
+- \usepackage{etoolbox}
+- \AtBeginEnvironment{verbatim}{\fontsize{7}{7}\selectfont}  # This sets the font size for verbatim environments
+#- \input{beamer_theme_trr266.sty}
 bibliography: references.bib
 biblio-style: apsr
-format:
-  pdf:
-    documentclass: article
-    number-sections: true
-    toc: false
-fig_caption: yes
-fontsize: 11pt
-ident: yes
-always_allow_html: yes
-header-includes:
-  - \usepackage[nolists]{endfloat}    
-  - \usepackage{setspace}\doublespacing
-  - \setlength{\parindent}{4em}
-  - \setlength{\parskip}{0em}
-  - \usepackage[hang,flushmargin]{footmisc}
-  - \usepackage{caption} 
-  - \captionsetup[table]{skip=24pt,font=bf}
-  - \usepackage{array}
-  - \usepackage{threeparttable}
-  - \usepackage{adjustbox}
-  - \usepackage{graphicx}
-  - \usepackage{csquotes}
-  - \usepackage{indentfirst}  # Added this line to ensure the first paragraph is indented
-  - \usepackage[margin=1in]{geometry}
 ---
 
-\pagebreak
+# Motivation
 
-# List of Abbreviations
-\setlength{\parindent}{0em}
+## Understanding Residual Income and Importance of P/B Ratios
 
-**AMEX**: American Stock Exchange  
-**CAPM**: Capital Asset Pricing Model  
-**BVE**: Book Value of Equity  
-**IDE**: Integrated Development Environment  
-**MVE**: Market Value of Equity  
-**NYSE**: New York Stock Exchange  
-**P/B**: Price-to-Book  
-**UK**: United Kingdom  
-**US**: United States
+- Residual income is a key metric in evaluating the performance of public firms.
+- It helps in assessing the value creation for shareholders beyond the cost of capital.
+- Price-to-Book (P/B) ratios are widely used in financial analysis to compare a firm’s market value to its book value.
+- Grouping firms by P/B ratios allows for a detailed analysis of financial performance across different valuation levels.
 
-\setlength{\parindent}{4em}
+## Project Objective
 
-\pagebreak
+- The primary objective is to demonstrate a reproducible and collaborative research workflow by calculating and analyzing residual incomes and Price-to-Book (P/B) ratios. 
+- The study aims to explore the residual income of U.S. public firms grouped by P/B ratios using external Worldscope datasets for the time range 1996 to 2015.
+- The project replicates the study of @Penman_2013 that links current P/B ratios to future residual income.
+- The analysis provides insights into the median residual income over time across different P/B groups, similar to the insights of Penman (2013).
 
-# Introduction
+## Project Relevance
 
-The primary objective is to demonstrate a reproducible and collaborative research workflow by calculating and analyzing residual incomes and Price-to-Book (P/B) ratios.
-The analysis includes filtering relevant data, identifying key financial years, grouping firms based on their P/B ratios, and calculating delfated residual incomes for each firm/year.
-
-Assignment I aims to replicate an empirical table that involves calculating the residual income of US firms over a defined period and examining the relationship between residual income and the Price-to-Book (P/B) ratio.
-The replication process includes data preparation, cleaning, normalization, and the application of statistical methods to compute and interpret financial metrics using Python.
-By following a step-by-step approach, I provide insights into the trends of residual earnings across different P/B groups, identical to trends presented by @Penman_2013.
-This document presents the analysis performed in Assignment I, adapted to the open science workflow.
-Note that intermediate check steps have been omitted for brevity.
-The full code, including intermediate print steps to verify outputs, is presented in `code/python/Assignment1_analyze_residual_income.py`.
+- Understanding the distribution and trends of residual income can inform investment decisions and corporate strategies.
+- It supports the broader objective of promoting transparency and reproducibility in empirical accounting research.
 
 # Assumptions and Notes
-
 The aim of Assignment I is to replicate a specific empirical table that involves calculating the residual income of US firms over a defined period and examining the relationship between residual income and the Price-to-Book (P/B) ratio.
 The replication process involves data preparation, cleaning, and normalization, followed by the application of statistical methods to compute and interpret financial metrics.
+
 For Assignment I, I used the Python programming language to carry out the empirical analysis.
@@
 ## Assumptions
-1.
-A constant cost of equity is assumed to simplify the calculation of residual income across multiple years, making the analysis more straightforward and easier to interpret.
-This approach allows for a direct comparison of performance across firms and time periods without the increased complexity due to fluctuating rates.
-In contrary, estimating a dynamic cost of equity typically involves financial models such as the Capital Asset Pricing Model (CAPM), which would require additional data like risk-free rates and firm beta values that are not available.
-2.
-The level of cost of equity capital is assumed to be 8% (re = 0.08), a reasonable average estimate for the period 1996-2015 in the US (@Goedhart_Koller_Williams_2002; @Damodaran_2024).
-This is assumption reflects the return equity investors typically expect given the level of risk during that period.
-3. The entire Worldscope panel data provides the Book Value of Equity (BVE), Market Value of Equity (MVE), and net income as values for the end of the respective fiscal year.
-4.
-The beginning of Year 0 data is equivalent to the end of the previous year because the financial data recorded at the end of one year represents the starting values for the following year.
-Therefore, residual income is deflated by the BVE at the end of the year before Year 0 to ensure comparability across different firms and time periods.
-5.
-It is assumed that Penman (2013) restricted the possible P/B values to a range of 0 to 7 to exclude outliers.
-Hence, extreme P/B values that are negative and very high (greater than 7) are excluded to focus the analysis on firms with more stable and reasonable valuations, reducing the impact of outliers.
+- **Constant Cost of Equity:** Assumed at 8% ($r_e = 8\%$), a reasonable average estimate for the period 1996-2015 in the US (@Goedhart_Koller_Williams_2002; @Damodaran_2024).
+- **Data Source:** Worldscope panel data includes Book Value of Equity (BVE), Market Value of Equity (MVE), and net income at fiscal year-end .
+- **Year 0 Data:** Year-end data from the previous year used as the starting values for the following year.
+- **Deflation Basis:** Residual income deflated by BVE at the end of the year before Year 0 for comparability.
+- **P/B Values Range:** Restricted to 0-7 to exclude outliers, focusing on firms with stable valuations.
 
-By following the steps provided in Section 4 and adhering to the assumptions made, I successfully replicated the analysis and produced the required table.
-A thorough step-by-step approach helped to understand and verify the outputs.
+# Data - Replication Steps
 
-# Replication Steps
-## Step 1: Data Loading and Merging
-Begin by loading the data from the provided Worldscope files.
-The static data contains information on unique ISIN IDs for each firm, as well as the corresponding firm name and country.
-The panel data provides the financial performance numbers of each ISIN.
-Merge the data on the ‘isin’ column to include country information in the panel data.
-This merging step is crucial for filtering the US firms specifically in further steps.
+Step 1: Data Loading and Merging
+
+Step 2: Filter Data for US Firms and Relevant Years
+
+Step 3: Identify Year 0 to Year 6 for Each Firm
+
+Step 4: Calculate P/B Ratio for Each Firm in Year 0
+
++ Use the following formula to calculate the P/B ratio:
+$$
+P/B_t = \frac{\textit{Market value of equity}_t}{\textit{Book value of equity}_t}
+$$
+
+Step 5: Form P/B Groups
+
+---
+
+Step 6: Calculate Residual Income
+
++ The Worldscope panel data provides BVE, MVE, and net income values at the end of each fiscal year, confirmed by the *2020 Annual Report* (2021) of ISIN AN8068571086 [@Schlumberger_2021, p. 20].
++ Calculate the non-deflated residual income for each firm/year using the following formula:
+$$
+\text{Residual income}_t = \text{Net income}_t - r_e \times \text{Book value of equity}_{t-1}
+$$
+
++ Calculate the deflated residual income for each firm/year using the following general formula:
+$$
+\text{Deflated residual income}_t = \frac{\text{Residual income}_t}{\text{Book value of equity}_{\text{Year 0 start}}}
+$$
+
+---
+
+Step 7: Output the Replicated Table
 
 ```{python}
-#| label: step1
-#| echo: true
+#| label: step1-7
+#| echo: false
 #| output: true
@@
 )
-```
-
-## Step 2: Filter Data for US Firms and Relevant Years
-Filter the merged data to include only US firms (excluding UK and German firms) and the year range 1996-2015.
-This ensures that further calculations are based on the specific subset of data that matches the assignment’s criteria.
-```{python}
-#| label: step2
-#| echo: true
-#| output: true
 
@@
 ].copy()
-```
-
-## Step 3: Identify Year 0 to Year 6 for Each Firm
-To calculate the P/B values and residual incomes for each firm/year, it is necessary to identify the base year (Year 0) and the subsequent relative years within the range 0 – 6 in the filtered data, as presented by Penman (2013).
-Define Year 0 as the earliest year in the dataset for each firm and display it in a new column ‘year_0’ that holds this value, making ‘year_0’ the same for each ISIN.
-Additionally, create the column ‘relative_year’, represented as the difference between the current year and Year 0.
-Since a firm may have more than 7 years of observations in the filtered data, limit the data to include only the first seven years (Year 0 to Year 6) for each ISIN to match the assignment’s requirements.
-Notably, the number of observations after merging Step 1 (298,120), filtering Step 2 (142,837), and identifying years in Step 3 (89,584) nearly halves with each step.
-
-```{python}
-#| label: step3
-#| echo: true
-#| output: true
 
@@
 filtered_data = filtered_data[filtered_data['relative_year'] <= 6]
-```
-
-## Step 4: Calculate P/B Ratio for Each Firm in Year 0
-According to Penman (2013), the residual earnings are calculated after P/B groups are formed in Year 0.
-Therefore, consider the firms’ financial performance data only for Year 0 to calculate the P/B ratio.
-Replace zero and infinite BVE values with NaN to drop invalid calculations
-(affecting 28 observations in total). Use the following formula to calculate the P/B ratio:
-$$
-P/B_t = \frac{\textit{Market value of equity}_t}{\textit{Book value of equity}_t}
-$$
-
-```{python}
-#| label: step4
-#| echo: true
-#| output: true
 
@@
 year_0_data = year_0_data.dropna(subset=['p/b'])
-```
-
-## Step 5: Form P/B Groups
-Sort the firms based on their P/B ratios in descending order and divide them into 20 P/B groups, with the highest P/B ratios in group 1 and the lowest in group 20, as presented by Penman (2013).
-As a result, each P/B group contains approximately 900 firms, ensuring an equal division of the data set.
-Calculate the median P/B value for each group and merge the P/B group assignment information back into the filtered data, providing information on which P/B group each firm belongs to in each year.
-```{python}
-#| label: step5
-#| echo: true
-#| output: true
 
@@
 filtered_data['p/b_group'] = filtered_data.groupby('isin')['p/b_group'].ffill()
-```
-
-## Step 6: Calculate Residual Income
-This step involves calculating the residual income for each P/B group for Years 0 to 6, followed by deflation by the BVE at the beginning of Year 0.
-Assume a cost of equity capital of 8%, a reasonable assumption for US firms in the given time range.
-Based on Penman (2013), the residual income for each company/year should be deflated by the book value of equity at the beginning of Year 0.
-
-To identify whether the financial data from Worldscope contains the numbers as of the beginning or end of the fiscal year, a random sample check was conducted.
-For example, for ISIN AN8068571086 in its *2020 Annual Report* (2021), the values for total assets, stockholders’ equity, net income, and other financial metrics are measured at the end of each fiscal year [@Schlumberger_2021, p.
-20].
-Hence, assume that the entire Worldscope panel data provides the BVE, MVE, and net income as values for the end of the respective fiscal year.
-
-In case a company has observations before 1996, the filtered data set does not enable calculation of residual income in Year 0 (1996) due to missing value of BVE in Year -1, resulting into NaN.
-To avoid that, create a copy of the merged data from Step 1 to preserve all years of observations.
-Filter it to US companies and identify the year_minus_1 for each firm.
-Extract BVE for the year before Year 0 (relative_year = -1) for the deflation step and extract the BVE of the previous year for each year for the residual income calculation step.
-
-Calculate the non-deflated residual income for each firm/year using the following formula:
-$$
-\text{Residual income}_t = \text{Net income}_t - r_e \times \text{Book value of equity}_{t-1}
-$$
-
-Assume that the beginning of Year 0 is equivalent to the end of the previous year because the financial data recorded at the end of one year represents the starting values for the following year.
-According to Penman (2013), residual income is deflated by book value at the beginning of year 0.
-Hence, for deflating the residual income, use the book value of equity from the year before Year 0 (bve_deflation) for all years (Years 1 to 6).
-This means that the BVE used for deflation remains constant across all years, normalizing the residual income across different firms and time periods.
-For example, for Year 0 (1996), use the BVE at the end of 1995 to deflate the residual income.
-
-Calculate the deflated residual income for each firm/year using the following general formula:
-
-$$
-\text{Deflated residual income}_t = \frac{\text{Residual income}_t}{\text{Book value of equity}_{\text{Year 0 start}}}
-$$
-
-```{python}
-#| label: step6
-#| echo: true
-#| output: true
 
@@
 )
-```
-
-## Step 7: Output the Replicated Table
-Calculate the median deflated residual income for each P/B group and relative year.
-Include the median P/B values in the output table.
-Ensure consistency with Penman (2013) by formatting the results to three decimal places for residual income and two decimal places for P/B values.
-The resulting table will be saved into an Excel file in the `data/generated` folder of the repository.
-
-The output table certainly indicates a wide range of P/B ratios among the firms, with some extreme values in high and low P/B groups.
-There is a noticeable trend where firms with very high or very low P/B ratios have less stable and more extreme residual earnings over time, while middle P/B groups seem to show more stable trends as presented by Penman (2013), suggesting a balanced reflection of expectations between market valuation and expected financial performance.
-
-The variation in trends can be explained by the difference in data sets and possible additional assumptions made by Penman (2013) that are not available.
-For example, the assignment data set includes all USA firms from the Worldscope database, providing a broader sample than Penman’s original study, which included only NYSE and AMEX firms.
-Furthermore, the assignment analysis focuses on the years 1996 to 2015, a more recent period with different economic conditions compared to the time frame used in Penman’s (2013) study.
-```{python}
-#| label: step7
-#| echo: true
-#| output: true
 
@@
 
-## Step 8: Modification - Exclude Extreme P/B Values
-As indicated in Step 7, the output table certainly includes outliers.
-To replicate the more stable P/B values observed by Penman (2013), exclude extreme P/B values that are negative or higher than 7.
-Recalculate the P/B groups based on this filtered data.
-Merge the updated P/B group information back into the filtered data and repeat the calculation of median deflated residual income for each P/B group and relative year.
-Format the updated results with the required decimal places and save to an Excel file.
-The modified output of the replicated table, closely resembling Penman (2013), is presented in Table 2.
+---
 
-The modified table resembles more the one of Penman (2013) by presenting similar trends.
-In particular, the exclusion of extreme P/B values results in more stable and consistent residual earnings across P/B groups and years.
-High P/B groups show higher residual earnings initially, which slowly decline over time, and low P/B groups have negative residual earnings that tend to fluctuate but still remain negative.
+Step 8: Modification - Exclude Extreme P/B Values
+
 ```{python}
 #| label: step8
-#| echo: true
+#| echo: false
 #| output: true
@@
 # Conclusion
-This project demonstrates the effectiveness of using an open science and collaborative workflow for analyzing residual incomes and Price-to-Book ratios of U.S.
-public firms.
-By following a step-by-step approach and using the TRR 266 Template for Reproducible Empirical Accounting Research, I was able to replicate an empirical table from @Penman_2013 and provide insights into the trends of residual earnings across different P/B groups, similar to @Penman_2013.
 
-The analysis revealed that high P/B groups initially exhibit higher residual earnings, which slowly decline over time, while low P/B groups consistently show negative residual earnings.
-Excluding extreme P/B values resulted in more stable and consistent residual earnings across groups and years.
+## Summary of Findings
 
-In the future, this repository can be cloned or forked (if made public) to kickstart further projects on residual income analysis. Thanks for reading and enjoy!
+- Replicated an empirical table from Penman (2013), showcasing similar trends.
+- The project structure demonstrated the effectiveness of an open science and collaborative workflow.
+- This repository can be cloned or forked for further projects on residual income analysis.
 
-\pagebreak
+## Key Insights
 
-\setcounter{table}{0}
-\renewcommand{\thetable}{\arabic{table}}
+- From analysis of residual incomes and Price-to-Book (P/B) ratios of U.S. public firms: 
+    - High P/B groups initially exhibit higher residual earnings, which decline over time. 
+    - Low P/B groups consistently show negative residual earnings.
+- Residual income analysis by P/B ratios offers valuable insights that can promote better investment and corporate strategy decisions.
 

### 2024-07-11 doc/paper_mine.qmd
Created documentation presentation
Added cross-referencing in paper_mine.qmd

@@
 Ensure consistency with Penman (2013) by formatting the results to three decimal places for residual income and two decimal places for P/B values.
-The resulting table will be saved into an Excel file in the `data/generated` folder of the repository.
+The resulting table of the Python code will be saved into an Excel file in the `data/generated` folder of the repository.
 
-The output table certainly indicates a wide range of P/B ratios among the firms, with some extreme values in high and low P/B groups.
+The output @tbl-median-residual-income certainly indicates a wide range of P/B ratios among the firms, with some extreme values in high and low P/B groups.
 There is a noticeable trend where firms with very high or very low P/B ratios have less stable and more extreme residual earnings over time, while middle P/B groups seem to show more stable trends as presented by Penman (2013), suggesting a balanced reflection of expectations between market valuation and expected financial performance.
@@
 ```{python}
-#| label: step7
+#| label: tbl-median-residual-income
+#| tbl-cap: "Median Residual Income by P/B Group and Relative Year"
 #| echo: true
-#| output: true
+#| output: asis
 
@@
 # Display the median residual income table 
-print(
-    "Median Residual Income by P/B Group and Relative Year:\n", 
-    median_residual_income
-)
+html_table = median_residual_income.to_html(classes='dataframe', border=0)
+html_content = f'<div style="max-width: 80%; margin: auto;">{html_table}</div>'
+from IPython.display import display, HTML
+display(HTML(html_content))
 ```
@@
 ## Step 8: Modification - Exclude Extreme P/B Values
-As indicated in Step 7, the output table certainly includes outliers.
+As indicated in Step 7, the output @tbl-median-residual-income certainly includes outliers.
 To replicate the more stable P/B values observed by Penman (2013), exclude extreme P/B values that are negative or higher than 7.
@@
 Format the updated results with the required decimal places and save to an Excel file.
-The modified output of the replicated table, closely resembling Penman (2013), is presented in Table 2.
+The modified output of the replicated @tbl-median-residual-income, closely resembling Penman (2013), is presented in @tbl-median-residual-income-filtered.
 
-The modified table resembles more the one of Penman (2013) by presenting similar trends.
+The modified @tbl-median-residual-income-filtered resembles more the one of Penman (2013) by presenting similar trends.
 In particular, the exclusion of extreme P/B values results in more stable and consistent residual earnings across P/B groups and years.
@@
 ```{python}
-#| label: step8
+#| label: tbl-median-residual-income-filtered
+#| tbl-cap: "Median Residual Income by P/B Group and Relative Year after Excluding Extreme P/B Values"
 #| echo: true
-#| output: true
+#| output: asis
 
@@
 # Display the median residual income table
-print(
-    "Median Residual Income by P/B Group and Relative Year:\n",
-    median_residual_income_filtered
-)
+html_table_filtered = median_residual_income_filtered.to_html(classes='dataframe', border=0)
+html_content_filtered = f'<div style="max-width: 80%; margin: auto;">{html_table_filtered}</div>'
+from IPython.display import display, HTML
+display(HTML(html_content_filtered))
 ```

### 2024-07-11 doc/presentation_mine.qmd
- Converted trr266_logo.eps to trr266_logo.pdf using Ghostscript for compatibility with XeLaTeX. - Updated beamer_theme_trr266.sty to reference the new PDF logo file.

@@
 - \AtBeginEnvironment{verbatim}{\fontsize{7}{7}\selectfont}  # This sets the font size for verbatim environments
-#- \input{beamer_theme_trr266.sty}
+- \input{beamer_theme_trr266.sty}
 bibliography: references.bib

### 2024-07-12 doc/paper_mine.qmd
final commit

@@
 # Display the median residual income table
-html_table_filtered = median_residual_income_filtered.to_html(classes='dataframe', border=0)
-html_content_filtered = f'<div style="max-width: 80%; margin: auto;">{html_table_filtered}</div>'
+html_table_filtered = median_residual_income_filtered.to_html(
+    classes='dataframe', 
+    border=0
+)
+
+html_content_filtered = (
+    f'<div style="max-width: 80%; margin: auto;">'
+    f'{html_table_filtered}'
+    f'</div>'
+)
 from IPython.display import display, HTML

### 2024-07-12 doc/paper_mine.qmd
Final commit

@@
 
-Assignment I aims to replicate an empirical table that involves calculating the residual income of US firms over a defined period and examining the relationship between residual income and the Price-to-Book (P/B) ratio.
+Assignment I aims to replicate an empirical table that involves calculating the residual income of US firms over a defined period and examining the relationship between residual income and the P/B ratio.
 The replication process includes data preparation, cleaning, normalization, and the application of statistical methods to compute and interpret financial metrics using Python.
@@
 
-The aim of Assignment I is to replicate a specific empirical table that involves calculating the residual income of US firms over a defined period and examining the relationship between residual income and the Price-to-Book (P/B) ratio.
+The aim of Assignment I is to replicate a specific empirical table that involves calculating the residual income of US firms over a defined period and examining the relationship between residual income and the P/B ratio.
 The replication process involves data preparation, cleaning, and normalization, followed by the application of statistical methods to compute and interpret financial metrics.
@@
 
-In the future, this repository can be cloned or forked (if made public) to kickstart further projects on residual income analysis. Thanks for reading and enjoy!
+In the future, this repository can be cloned or forked (if made public) to kickstart further projects on residual income analysis. Thanks for reading!
 
